---
title: "RNA-seq Data"
format: docx
editor: visual
---

```{r}
#| inclue: false
#| message: false

library(readr)
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())
library(dplyr)
library(tidyr)
library(hexbin)


source('https://tinyurl.com/wjm6ryj')

if(FALSE)
{
    download.file(url = "https://github.com/Bioconductor/bioconductor-teaching/raw/master/data/GSE96870/rnaseq.csv",
              destfile = "data/rnaseq.csv")
}

### Read in data here ###
dat <- read_csv("data/rnaseq.csv") %>%
  mutate(ln2expression = log2(expression +1))
```

Note: this lab borrows heavily from a [Carpentries incubator module on visualization](https://carpentries-incubator.github.io/bioc-intro/40-visualization/index.html).

## RNA-Seq data

To start out with, let's explore the data provided:

```{r}

#quick view of the data
dat

#organisms
table(dat$organism)

#summary
str(dat)

#gene_biotype
table(dat$gene_biotype)

#number of genes
str(unique(dat$gene))

#infection status
str(unique(dat$infection))

#sex
table(dat$sex)

```

Take a look at the distribution of the expression levels:

```{r}
#| message: false
# hint: use geom_histogram
## You'll probably notice an automatic message that appears when drawing the histogram: `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
## Change the arguments bins or binwidth of geom_histogram() to change the number or width of the bins.

(g1 <- ggplot(dat, aes(expression)) +
  geom_histogram(bins = 50))
  
```

The data should be transformed. Transforming on the $\log_{10}$ scale is easy in `ggplot2`,

```{r}
#| warning: false
# transform x-axis to log_10 scale

(g2 <- g1 +
  scale_x_log10())
```

but we often use a $log_2$ scale when looking at expression data:

```{r}
#| warning: false

# the following code will be helpful
ticks_at <- c(0,5,10,15)
    
ggplot(dat, aes(ln2expression)) +
  geom_histogram(bins=50) +
  scale_x_continuous(breaks = ticks_at, labels = 2^ticks_at) +
  labs(x = "Expression")
      
      
```

### Building plots iteratively

We will now draw a scatter plot with two continuous variables and the `geom_point()` function. This graph will represent the $\log_2$ fold changes of expression comparing time 8 versus time 0, and time 4 versus time 0. To this end, we first need to compute the means of the log-transformed expression values by gene and time, then the log fold changes by subtracting the mean log expressions between time 8 and time 0 and between time 4 and time 0.

```{r}
#| message: false
# fold change data
dat_fc <- dat %>%
  

     # calculate summaries of log2 expression grouped by gene, time, gene_biotype
  group_by(gene, time, gene_biotype) %>%
  summarize(mean_exp = mean(ln2expression)) %>%
  ungroup() %>%

    # pivot wider so that there is a separate column for each value in time
  pivot_wider(names_from = time, values_from = mean_exp) %>%
  mutate(time_8_vs_0 = `8` - `0`, # compare time 8 vs 0
          time_4_vs_0 = `4` - `0`) # compare time 4 vs 0

# visualize time_4_vs_0 by time_8_vs_0 
ggplot(dat_fc, aes(time_4_vs_0, time_8_vs_0)) +
  geom_point(alpha = 0.1) +
# hint: the following line will be helpful
    geom_abline(intercept = 0, slope = 1)
```

We could also add color, but it begins to get a little messy with the large number of categories we have.

```{r}
ggplot(dat_fc, aes(time_4_vs_0, time_8_vs_0, color = gene_biotype)) +
  geom_point(alpha = 0.1) +
  geom_abline(intercept = 0, slope = 1) +
  scale_color_manual(values = c(cbpalette, "#555555"))
```

Over-plotting can be an issue with large datasets like this. One solution is using `hexbin:geom_hex()`.

```{r}
ggplot(dat_fc, aes(time_4_vs_0, time_8_vs_0)) +
  geom_abline(intercept = 0, slope = 1) +
  geom_hex()
```

### Box plots

We can use boxplots to visualize the distribution of gene expressions within each sample:

```{r}
ggplot(dat, aes(sample, ln2expression)) +
  geom_boxplot() + 
  labs(y = "Log2 Expression") +
  theme(axis.text.x = element_text(angle = 90,  hjust = 0.5, vjust = 0.5))
```

Adding points to box plots can give us a better understanding of the underlying distributions.

```{r}
# hints: use geom_jitter to plot points; this line of code will be helpful
ggplot(dat, aes(sample, ln2expression)) +
  geom_jitter(width = 0.1, alpha = 0.03, color = cbbPalette[2]) +
  geom_boxplot(alpha = 0.0) + 
  labs(y = "Log2 Expression") +
  theme(axis.text.x = element_text(angle = 90,  hjust = 0.5, vjust = 0.5)) 
```

We could also use `geom_violin` to better see the distribution of points. Try coloring the points by `time` as well.

```{r}
mutate(dat, time = as.character(time)) %>%
ggplot(aes(sample, ln2expression)) +
  geom_jitter(aes(color = time), width = 0.1, alpha = 0.03) +
  geom_violin(alpha = 0) + 
  labs(y = "Log2 Expression") +
  theme(axis.text.x = element_text(angle = 90,  hjust = 0.5, vjust = 0.5)) +
  scale_color_manual(values = cbbPalette[1:3]) +
  guides(color = guide_legend(override.aes = list(alpha = 1)))
```

### Line plots

Let's calculate the mean expression per duration of the infection for the 10 genes having the highest log fold changes comparing time 8 versus time 0. We can do this either by coloring each gene or using `facet_wrap`.

```{r}
#| message: false

# First, we need to select the top 10 genes
top_10_gene <- arrange(dat_fc, desc(abs(time_8_vs_0)))$gene[1:10]
# Then create a subset of dat containing the 10 selected genes
top_10_df <- filter(dat, gene %in% top_10_gene) %>%
  group_by(gene, time) %>%
  summarize(mean_exp = mean(ln2expression)) %>%
  ungroup()

    # Then group the data by gene and time
    # calculate the mean gene expression within each group

# plot summary lines
ggplot(top_10_df, aes(time, mean_exp, group = gene, color = gene)) +
  geom_line() +
  scale_color_manual(values = cbpalette) +
  labs(y="mean log2 expression")


#with facet wrap
ggplot(top_10_df, aes(time, mean_exp)) +
  geom_line() +
  labs(y="mean log2 expression") +
  facet_wrap(~gene)
```

Now we would like to split the line in each plot by the sex of the mice.

```{r}
#| message: false
# Create a subset of dat containing the 10 selected genes
top_10_df2 <- filter(dat, gene %in% top_10_gene) %>%
  group_by(gene, time, sex) %>%
  summarize(mean_exp = mean(ln2expression)) %>%
  ungroup()


#with facet wrap
ggplot(top_10_df2, aes(time, mean_exp, group = sex, color = sex)) +
  geom_line() +
  labs(y="mean log2 expression") +
  facet_wrap(~gene) +
  scale_color_manual(values = cbbPalette)

```

Let's do something similar and create a plot that depicts how the average expression of each chromosome changes through the duration of infection.

```{r}
#| message: false

#summarize over chromosomes
chr_df3 <- dat %>%
  mutate(chromosome_name = factor(chromosome_name, levels = c(1:19, 'X', 'Y')))%>%
  group_by(chromosome_name, time) %>%
  summarize(mean_exp = mean(ln2expression)) %>%
  ungroup()

ggplot(chr_df3, aes(time, mean_exp)) +
  geom_line() + 
  facet_wrap(~chromosome_name, scales = "free_y") 
  
```

The `facet_wrap` geometry extracts plots into an arbitrary number of dimensions to allow them to cleanly fit on one page. On the other hand, the `facet_grid` geometry allows you to explicitly specify how you want your plots to be arranged via formula notation (`rows ~ columns`; a `.` can be used as a placeholder that indicates only one row or column).

Let's modify the previous plot to compare how the mean gene expression of males and females has changed through time:

```{r}
# One column, facet by rows
ggplot(top_10_df2, aes(time, mean_exp, group = gene, color = gene)) +
  geom_line() +
  scale_color_manual(values = cbpalette) +
  labs(y="mean log2 expression") +
  facet_grid(sex~.)
```
